Automatic Disease Diagnosis Using Large Language Models and Answer Set Programming

### 2025-06-08 paper/draft.tex
Nits

\maketitle
\begin{abstract}
Accurate disease prediction is vital for timely intervention, 
effective treatment, and reducing medical complications. 
While symbolic AI has been applied in healthcare, 
its adoption remains limited due to the complexity 
of required knowledge bases.
This work introduces \sys, a framework that combines Large Language Models (LLMs) 
with Answer Set Programming (ASP) to overcome this barrier. 
LLMs extract medical knowledge from text and translate it into ASP code, 
while ASP enables structured, explainable reasoning. 
This integration yields a robust, interpretable prediction framework 
that leverages the strengths of both paradigms. 
This approach shows strong performance on real patient data.
% \sys accurately predicts diseases based on symptoms derived from medical literature.
\end{abstract}

@@

The development of automatic disease prediction systems 
is crucial for enhancing healthcare and supporting medical 
professionals in timely and accurate diagnosis. 
Symbolic artificial intelligence (AI), particularly Answer Set Programming (ASP), 
has been explored in this context due to its ability 
to support structured and explainable reasoning. 
However, the construction of expressive knowledge bases 
remains a significant barrier to practical adoption.

@@

The central idea of Answer Set Programming is to represent a problem 
declaratively using logical rules, facts, and constraints. 
These elements collectively form a program that describes the conditions any solution must satisfy. 
Instead of writing an algorithm to solve the problem, 
the programmer encodes the specification of valid solutions. 
An ASP solver then processes this program to compute one or more stable models (also known as answer sets),
each of which corresponds to a solution that satisfies all the given rules and constraints.

In traditional programming, transitioning from a problem to a solution 
requires the programmer's thorough understanding of the given problem, 
followed by the creation of a program that will produce the correct output 
when provided with an instance of the problem, which will then be interpreted as the solution. 
In Answer Set Programming, the process of deriving solutions from a problem specification 
typically follows a sequence of well-defined steps \cite{Gebser2013}, 
as illustrated in \cref{fig:asp-solving}.

@@

\begin{itemize}
    \item \textbf{Modeling}: The problem is modeled using ASP (Answer Set Programming) language.
    \item \textbf{Grounding}: A grounder (e.g., gringo) converts the program into a set of basic rules and facts.
    \item \textbf{Solving}: A solver (e.g., clasp) finds a solution to the problem by computing the set of stable models (answer sets).
\end{itemize}

@@

\section{Related work}
% we need to add something about LLM + ASP in disease diagnosis
\paragraph{\textbf{ASP for disease diagnosis}}
Answer Set Programming (ASP) has been used to represent 
and reason over medical knowledge in a declarative, logic-based manner. 
Early work applied ASP to biomedical ontologies and structured databases 
to answer complex diagnostic and pharmacological queries with interpretable 
logical justifications \cite{Erdem2011}. In the clinical domain, 
ASP is especially effective for encoding medical rules and patient-specific data, 
enabling precise and explainable decision-making.
Common strategies in ASP-based medical reasoning include:

>>>>

\maketitle
\begin{abstract}
Accurate disease prediction is vital for timely intervention, 
effective treatment, and reducing medical complications. 
While symbolic AI has been applied in healthcare, 
its adoption remains limited due to the complexity 
of required knowledge bases.
This work introduces \sys, a framework that combines Large Language Models (LLMs) 
with Answer Set Programming (ASP) to overcome this barrier. 
LLMs extract medical knowledge from text and translate it into ASP code, 
while ASP enables structured, explainable reasoning. 
This integration yields a robust, interpretable prediction framework 
that leverages the strengths of both paradigms. 
\sys shows strong performance on real patient data.
% \sys accurately predicts diseases based on symptoms derived from medical literature.
\end{abstract}

@@

The development of automatic disease prediction systems 
is crucial for enhancing healthcare and supporting medical 
professionals in timely and accurate diagnosis. 
Symbolic artificial intelligence (AI), particularly ASP, 
has been explored in this context due to its ability 
to support structured and explainable reasoning. 
However, the construction of expressive knowledge bases 
remains a significant barrier to practical adoption.

@@

The central idea of ASP is to represent a problem 
declaratively using logical rules, facts, and constraints. 
These elements collectively form a program that describes the conditions any solution must satisfy. 
Instead of writing an algorithm to solve the problem, 
the programmer encodes the specification of valid solutions. 
An ASP solver then processes this program to compute one or more stable models (also known as answer sets),
each of which corresponds to a solution that satisfies all the given rules and constraints.

In traditional programming, transitioning from a problem to a solution 
requires the programmer's thorough understanding of the given problem, 
followed by the creation of a program that will produce the correct output 
when provided with an instance of the problem, which will then be interpreted as the solution. 
In ASP, the process of deriving solutions from a problem specification 
typically follows a sequence of well-defined steps \cite{Gebser2013}, 
as illustrated in \cref{fig:asp-solving}.

@@

\begin{itemize}
    \item \textbf{Modeling}: The problem is modeled using ASP language.
    \item \textbf{Grounding}: A grounder (e.g., gringo) converts the program into a set of basic rules and facts.
    \item \textbf{Solving}: A solver (e.g., clasp) finds a solution to the problem by computing the set of stable models (answer sets).
\end{itemize}

@@

\section{Related work}
% we need to add something about LLM + ASP in disease diagnosis
\paragraph{\textbf{ASP for disease diagnosis}}
ASP has been used to represent 
and reason over medical knowledge in a declarative, logic-based manner. 
Early work applied ASP to biomedical ontologies and structured databases 
to answer complex diagnostic and pharmacological queries with interpretable 
logical justifications \cite{Erdem2011}. In the clinical domain, 
ASP is especially effective for encoding medical rules and patient-specific data, 
enabling precise and explainable decision-making.
Common strategies in ASP-based medical reasoning include:
